---
title: "Deaf-of-Deaf: Data Methods & Results"
author: "Adam C Sales"
date: '`r Sys.Date()`'
output: "word_document"
---

```{r init,include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(echo = FALSE,cache=TRUE)

source('generalCode/estimationFunctions.r')

pn = function(x) prettyNum(x,big.mark = ',')
```

```{r readInData,cache=TRUE,include=FALSE}
if(!is.element("dat",ls())){
  dat <- read_csv('deafHHdata.csv')
  names(dat) <- tolower(names(dat))
}
```

```{r descriptivesInclGQ,include=TRUE,results='markup'}
if(sum(dat$gq==3)>0){
  nTot <- nrow(dat)
  nDeafTot <- sum(dat$diffhear==2)
  nDeafKid <- sum(dat$diffhear[dat$age<17]==2)
  
  deafByGQ <- dat%>%
    mutate(gq=c('Vacant','Households','Households',
                'Institutions'	,
                'Other group quarters',
                'Households')[gq+1]
    )%>%
    group_by(gq)%>%
    summarize(ngqSamp=n(),
              ngqPop=sum(perwt),
              ndeafSamp=sum(diffhear==2),
              nDeafPop=sum(perwt[diffhear==2]),
              ndeafKidSamp=sum(diffhear[age<17]==2),
              ndeafKidPop=sum(perwt[diffhear==2&age<17])
    )%>%
    bind_rows(summarise(.,
                        across(where(is.numeric), sum),
                        across(where(is.character), ~"Total")))%>%
    bind_rows(summarise(filter(.,gq%in%c('Institutions','Other group quarters')),
                        across(where(is.numeric), sum),
                        across(where(is.character), ~"Total GQ")))
  
  kable(deafByGQ,format.args=list(big.mark=','))
  
  
  dat <- dat%>%
    filter(gq%in%c(1,2,5)) ### not in group quarters
  
}
```

```{r deafMissigness}
deafMiss <- xtabs(~diffhear+qdifhear,data=dat)%>%
  as.matrix()%>%
  cbind(.,./rowSums(.)*100)%>%
  `dimnames<-`(list(c('Hearing','Deaf'),c('Observed N','Imputed N','Observed %','Imputed %')))

kable(deafMiss)
  

```


## Data

- 1- year 2019 ACS data from IPUMS.org
- Extracted: 11/29/2020
- ACS: 
  - run by census bureau
  - samples $\approx$1\% of US households'
  - includes either household or person-level weights to account for:
    - unequal sampling probabilities (i.e. complex sampling design)
    - survey non-response 
- Sample Size (original)
  - total: `r pn(nTot)`
  - total deaf people: `r pn(nDeafTot)`
  - total deaf <17: `r pn(nDeafKid)`
- Exclude group quarters
  - Examples: Dorm room, prison, group home
  - Excluded sample: 
    - in GQ: `r pn(deafByGQ$ngqSamp[deafByGQ$gq=='Total GQ'])`
    - deaf: `r pn(deafByGQ$ngqSamp[deafByGQ$gq=='Total GQ'])`
    - deaf <17: `r pn(deafByGQ$ndeafKidSamp[deafByGQ$gq=='Total GQ'])`
  - Excluded population (estimated):
    - in GQ: `r pn(deafByGQ$ngqPop[deafByGQ$gq=='Total GQ'])`
    - deaf: `r pn(deafByGQ$ngqPop[deafByGQ$gq=='Total GQ'])`
    - deaf <17: `r pn(deafByGQ$ndeafKidPop[deafByGQ$gq=='Total GQ'])`
  - Deaf missigness
    - item-level: respondent left the item empty
    - imputed by census bureau
    - deafness imputed for roughly `r round(deafMiss['Deaf','Imputed %'])`% of people labeled as "deaf" and `r round(deafMiss['Hearing','Imputed %'])`% of people labeled as "hearing"
    
    


## Methodology

### Definitions:

  - "Deaf": answered "Yes" to question: "Is this person deaf or does he/she have serious difficulty hearing?"
  - "Child": under the age of 17
  
### Inferring relationships

```{r mompoprulePercents}
popDirect <- mean(between(dat$poprule[dat$poprule>0],10,19))
momDirect <- mean(between(dat$poprule[dat$poprule>0],10,19))
```

- Followed IPUMS

General idea: ACS collects one form from each household, filled out by a member of the household, called "Person 1," who responds to questions about each member of the household.
One such question is "How is this person related to Person 1?"--that is, Person 1 is asked to list their [nb: how do we do gender? this is the "singular their"] relationship to each other household member.
IPUMS infers family interrelationships within a household using responses to this question, in addition to supplementary information such as subjects' surnames (which IPUMS has access to, but which are not public) and the order in which they are listed in the response form. 
A full description of this process can be found in the [IPUMS User Guide (Chapter 5)](https://usa.ipums.org/usa/chapter5/chapter5.shtml).
Most relationships (approximately 
`r round(popDirect*100)`%
of inferred father-child and 
`r round(momDirect*100)`%
of inferred mother-child relationships) were inferred unambiguously, including cases in which Person 1 identifies themselves as the parent or child of someone else in the household. 
The remainder of relationships were inferred following a set of rules, that are themselves based on assumptions regarding typical household structures, surnames, and other factors.

```{r exampleFamily,results='markup'}
dat%>%
  filter(serial==2405 )%>%
  transmute(`Person #`=pernum,
            `Name*`=c('Ava','Benjamin','Charlotte','Delilah'),
            `Rel. to Person 1`=c('Self',
                                 'Spouse',
                                 'Child',
                                 'Child-in-law',
                                 'Parent',
                                 'Parent-in-Law',
                                 'Sibling',
                                 'Sibling-in-Law',
                                 'Grandchild')[relate],
            Age=age,
            `Sex`=c('Male','Female')[sex],
            `Mother #`=ifelse(momloc==0,'-',momloc),
            `Father #`=ifelse(poploc==0,'-',poploc)
            )%>%
  kable(align='c')
```

For instance, consider the household in Table ??. Person 1 ("Ava") identified Person 2 ("Benjamin") as her spouse, Person 3 ("Charlotte") as her child, and Person 4 ("Delila") as her grandchild^[These names are made-up for clarity of exposition; they were taken from the [list of most common names](https://www.ssa.gov/cgi-bin/popularnames.cgi) for babies born in 2020, provided by the US Office of Social Security.]. Since Ava and Benjamin are married, and Charlotte is Ava's daughter, IPUMS assumes that Benjamin is Charlotte's father. Since Delila is Ava's granddaughter, Charlotte is the only one of Ava's daughters present, and Charlotte is between 12 and 54 years older than Delila, IPUMS lists Charlotte as Delila's mother. 
This latter assumption need not always hold--if Ava has other children (who are not members of the same household, and hence not listed), one of them may be Delila's father or mother, in which case Charlotte would be Delila's aunt, rather than her mother. If Ava had other children in the household, there would have been further ambiguity as to Delila's parents; in that case, IPUMS would use surnames, ages, and the order in which Ava listed household members to guess at Delila's parentage. 
  
  
### Estimating Population Quantitites

We estimate population quantities--i.e., population totals and percentages--using the person-level weights provided with the ACS data.
These weights are intended to generalize estimates to the 2019 population of individuals across the fifty states and Washington, DC, and that is how we interpret our estimates. 
This interpretation relies on a number of statistical assumptions, including the validity of the procedures that the Census Bureau uses to adjust the sampling weights for survey non-response and impute missing item-level responses, as well as the procedures IPUM uses to identify parentage and other family relationships within a household. 
Violations of these assumptions are impossible to check with available data, and may lead to bias.
However, assumptions of this nature are unavoidable in survey research, and the ACS sample is arguably the highest-quality data available for our purposes. 

We estimate standard errors for population quantities using the 80 replication weights provided in the ACS sample.

### Computation and Replication
The analyses were peformed in R (CITE) and replication data and code (in Rmarkdown; CITE) are available at [](https://github.com/nationalDeafCenter/deaf-of-deaf).


## Results

```{r moresubsetting,include=FALSE}
dat <- dat%>%
    group_by(serial)%>%
      mutate(ndeaf=sum(diffhear==2,na.rm=TRUE))%>%
        filter(ndeaf>0)%>% ### only look at HH with at least 1 deaf person
          ungroup()
gc()
```




```{r estimation}
famDat <- dat%>%
  group_by(serial)%>%
    summarize(nfams=nfams[1],nmothers=nmothers[1],nfathers=nfathers[1],npop=sum(poploc>0),
              nmom=sum(momloc>0),ownchildren=any(nchild>0),nkid=sum(age<17),
              ndeafkid=sum(age<17&diffhear==2),ndeaf=sum(diffhear==2),nppl=n())




deafKids <- dat%>%
  group_by(serial)%>%
    arrange(pernum)%>%
      mutate(
        pn2=1:n(),
        hasDad=poploc>0,hasMom=momloc>0,
        hasParent=hasDad|hasMom,
        deafDad=ifelse(hasDad,diffhear[poploc]==2,FALSE),
        deafMom=ifelse(hasMom,diffhear[momloc]==2,FALSE),
        deafDad2 = ifelse(poploc2==0,FALSE,diffhear[poploc2]==2),
        deafMom2 = ifelse(momloc2==0,FALSE,diffhear[momloc2]==2),
        numParents=I(momloc>0)+I(poploc>0)+I(momloc2>0)+I(poploc2>0),
        deafParent=deafMom|deafDad|deafDad2|deafMom2,
        deafDad=deafDad|deafDad2,
        deafMom=deafMom|deafMom2,
        numAdults=sum(age>17),
        numDeafAdults=sum(age>17 & diffhear==2),
        anyDeafAdults=numDeafAdults>0,
        numDeafParents=deafDad+deafDad2+deafMom+deafMom2,
        twoDeafParents=numDeafParents>=2,
        numOtherDeafChildren=sum(age<18&diffhear==2)-1
         )

## just to check
mean(deafKids$pernum==deafKids$pn2)

deafKids <- filter(deafKids,diffhear==2,age<=17)

deafKids%>%group_by(age)%>%summarize(hasParent=mean(hasParent))%>%ggplot(aes(age,hasParent))+geom_point()+ylim(0,1)+geom_smooth()

### estimate the total # in a subgroup & the % of non-GQ deaf kids who are in subgroup
## rounding rules based on https://arxiv.org/ftp/arxiv/papers/1301/1301.1034.pdf
estSubst <- function(subst,name){
  N <- svTot(deafKids,subst,w1='perwt',wrep=paste0('repwtp',1:80))
  N <- prettyNum(c(round(N[1:2],-floor(log(N['se'],10))+1),N[3]),big.mark=',')
  per <- estSEstr(subst,w1='perwt',wrep=paste0('repwtp',1:80),sdat=deafKids)
  roundPer <- ifelse(per['se']==0,0,-floor(log(per['se']*100,10))+1)
  per <- c(
      formatC(round(per[c(1,2)]*100,roundPer),roundPer,format='f'),
      prettyNum(per[3],big.mark=',')
  )
  cbind(tibble(`living with...`=c(name,''),`N/%`=c('N','%')),rbind(N,per))
}

############################################################################################################
#### Questions
### from https://docs.google.com/document/d/1LbjhZl6ZdEJJMl7wfSLOhn7OYDXLNXvaV7RXCK5_oMo/edit

## Among deaf children (17 or younger) living in non-institutionalized/group quarters,

### living with deaf adults:
## how many of them live with deaf adults 18+?
## How many live with more than one deaf adult?
## How many live with at least one deaf parent?
## How many live with two deaf parents?

### living with (other) deaf children:
## how many of them live with other deaf children?
## How many with more than one other deaf child?
############################################################################################################

### total # of deaf children <18 (excl. GQ)
total <- round(svTot(deafKids,w1='perwt',wrep=paste0('repwtp',1:80)))


### proportions deaf kids
kidEst <- bind_rows(
    estSubst('age<18','TOTAL (non-GQ deaf children)'),
    estSubst('numDeafParents>0','at least 1 deaf parent'),
    estSubst('numDeafParents==1','exactly 1 deaf parent'),
    estSubst('numDeafParents==2','exactly 2 deaf parents'),
    estSubst('numDeafAdults>0','at least 1 deaf adult'),
    estSubst('numDeafAdults>1','at least 2 deaf adults'),
    estSubst('numOtherDeafChildren>0','at least 1 other deaf kid'),
    estSubst('numOtherDeafChildren>1','at least 2 other deaf kids'))%>%
    add_case(`living with...`="NOTES:")%>%
    add_case(`living with...`=paste("Year=",paste(unique(dat$year),collapse=', ')))%>%
    add_case(`living with...`="Excluding deaf children in group quarters")%>%
    add_case(`living with...`="Child: <18; Adult: 18+")%>%
    add_case(`living with...`="Denominator for % is all deaf children")

kidEst[is.na(kidEst)] <- ''
```

#write.csv(kidEst,'deafKidsWdeafParentsAndKids.csv',row.names=FALSE)







